---
title: "Venn Diagram"
output: html_document
---

```{r setup, include=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(clusterProfiler)
library(xtable)
library(writexl)
library(VennDiagram)
library(GO.db)
library(biomaRt)
```

All significantly regulated genes from the *GCGR Ab* and *GcgrKO* datasets are loaded:
```{r}
All_Sig_Genes_GCGR_Ab <- read.csv('~/All_Sig_Genes_GCGRAb_CtlAb.csv') 
All_Sig_Genes_KOvsWT <- read.csv('~/"All_Sig_Genes_KOvsWT.csv"')
```

Comparing genes from the two datasets and exporting a list of the overlapping genes: 
```{r message=FALSE, warning=FALSE}
KO_Data_Venn_Up <- All_Sig_Genes_KOvsWT %>% 
  filter(., log2FoldChange > 0) %>% 
  dplyr::select('ENSEMBL') %>%
  c(unlist(.['ENSEMBL'])) %>% .[-1]
KO_Data_Venn_Down <- All_Sig_Genes_KOvsWT %>% 
  filter(., log2FoldChange < 0) %>% 
  dplyr::select('ENSEMBL') %>%
  c(unlist(.['ENSEMBL'])) %>% .[-1]
GCGR_Ab_Data_Venn_Up <- All_Sig_Genes_GCGR_Ab %>% 
  filter(., log2FoldChange > 0) %>% 
  dplyr::select('ENSEMBL') %>%
  c(unlist(.['ENSEMBL'])) %>% .[-1]
GCGR_Ab_Data_Venn_Down <- All_Sig_Genes_GCGR_Ab %>% 
  filter(., log2FoldChange < 0) %>% 
  dplyr::select('ENSEMBL') %>%
  c(unlist(.['ENSEMBL'])) %>% .[-1]

venn.diagram(x = list(KO_Data_Venn_Up,GCGR_Ab_Data_Venn_Up),
             category.names = c(expression("GCGR"^"-/-") , "GCGR Ab"),
             filename = 'VennDiagram_GCGR_KO_Ab_UP.png',
             output=TRUE,
             main = "Up-regulated genes",
             main.pos = c(0.5,0.95),
             main.cex = 0.9,
          imagetype="png",
          resolution = 1000,
          compression = "lzw",
          lty = "blank",
          fill = c("#330099", "#33CC33"),
          cex = 0.8,
          fontfamily = "serif",
          cat.cex = 0.8,
          cat.pos = c(-0, 0),
          cat.dist = c(0.03, 0.03),
          cat.fontfamily = "serif",
          disable.logging = TRUE)

venn.diagram(x = list(KO_Data_Venn_Down,GCGR_Ab_Data_Venn_Down),
             category.names = c(expression("GCGR"^"-/-") , "GCGR Ab"),
             filename = 'VennDiagram_GCGR_KO_Ab_DOWN.png',
             output=TRUE,
             main = "Down-regulated genes",
             main.pos = c(0.5,0.95),
             main.cex = 0.9,
          imagetype="png",
          resolution = 1000,
          compression = "lzw",
          lty = "blank",
          fill = c("#330099", "#33CC33"),
          cex = 0.8,
          fontfamily = "serif",
          cat.cex = 0.8,
          cat.pos = c(-0, 0),
          cat.dist = c(0.03, 0.03),
          cat.fontfamily = "serif",
          disable.logging = TRUE)

Down_Overlap <- calculate.overlap(x = list(KO_Data_Venn_Down,GCGR_Ab_Data_Venn_Down)) %>% .[["a3"]] %>% as.data.frame(row.names = "ENSEMBL") %>% t() %>% as_tibble() %>% semi_join(All_Sig_Genes_GCGR_Ab,., by = "ENSEMBL") %>% dplyr::select("ENSEMBL","gene_name","gene_symbol","padj","foldChange")

Up_Overlap <- calculate.overlap(x = list(KO_Data_Venn_Up,GCGR_Ab_Data_Venn_Up)) %>% .[["a3"]] %>% as.data.frame(row.names = "ENSEMBL") %>% t() %>% as_tibble() %>% semi_join(All_Sig_Genes_GCGR_Ab,., by = "ENSEMBL") %>% dplyr::select("ENSEMBL","gene_name","gene_symbol", "padj","foldChange")

All_Sig_Genes_KOvsWT_Overlap <- rename(All_Sig_Genes_KOvsWT, foldChange_KO = foldChange, padj_KO = padj) %>% dplyr::select("ENSEMBL", "padj_KO","foldChange_KO")

Down_Overlap <- left_join(Down_Overlap, All_Sig_Genes_KOvsWT_Overlap, by = "ENSEMBL") %>% rename(., foldChange_Ab = foldChange, padj_Ab = padj)
Up_Overlap <- left_join(Up_Overlap, All_Sig_Genes_KOvsWT_Overlap, by = "ENSEMBL") %>% rename(., foldChange_Ab = foldChange, padj_Ab = padj)

write_xlsx(Down_Overlap, "Overlapping_Down_Genes_GcgrKO_GCGR-Ab.xlsx")
write_xlsx(Up_Overlap, "Overlapping_UP_Genes_GcgrKO_GCGR-Ab.xlsx")
```

Creating a list of lipid annotated genes (GOBP annotated): 
```{r message=FALSE, warning=FALSE}
LipidGOBP <- unique(c(GOBPOFFSPRING[["GO:0006629"]],GOBPOFFSPRING[["GO:0010876"]],GOBPOFFSPRING[["GO:0055088"]]))

Lipid_BP_Genes <- bitr(LipidGOBP, fromType="GOALL", toType= "SYMBOL", OrgDb='org.Mm.eg.db')
Lipid_BP_Genes <- tibble("gene_symbol" = unique(Lipid_BP_Genes$SYMBOL))
```
  
Comparing lipid related genes in the two datasets and exporting a list of the overlapping genes:
```{r eval=FALSE, include=TRUE}
KO_Data_Venn_Up <- All_Sig_Genes_KOvsWT %>% semi_join(., Lipid_BP_Genes, by = "gene_symbol") %>%
  filter(., log2FoldChange > 0) %>% 
  dplyr::select('ENSEMBL') %>%
  c(unlist(.['ENSEMBL'])) %>% .[-1]
KO_Data_Venn_Down <- All_Sig_Genes_KOvsWT %>% semi_join(., Lipid_BP_Genes, by = "gene_symbol") %>%
  filter(., log2FoldChange < 0) %>% 
  dplyr::select('ENSEMBL') %>%
  c(unlist(.['ENSEMBL'])) %>% .[-1]
GCGR_Ab_Data_Venn_Up <- All_Sig_Genes_GCGR_Ab %>% semi_join(., Lipid_BP_Genes, by = "gene_symbol") %>%
  filter(., log2FoldChange > 0) %>% 
  dplyr::select('ENSEMBL') %>%
  c(unlist(.['ENSEMBL'])) %>% .[-1]
GCGR_Ab_Data_Venn_Down <- All_Sig_Genes_GCGR_Ab %>% semi_join(., Lipid_BP_Genes, by = "gene_symbol") %>%
  filter(., log2FoldChange < 0) %>% 
  dplyr::select('ENSEMBL') %>%
  c(unlist(.['ENSEMBL'])) %>% .[-1]

venn.diagram(x = list(KO_Data_Venn_Up,GCGR_Ab_Data_Venn_Up),
             category.names = c(expression("GCGR"^"-/-") , "GCGR Ab"),
             filename = 'VennDiagram_GCGR_KO_Ab_UP_LIPID.png',
             output=TRUE,
             main = "Up-regulated genes",
             main.pos = c(0.5,0.95),
             main.cex = 0.9,
          imagetype="png",
          resolution = 1000,
          compression = "lzw",
          lty = "blank",
          fill = c("#330099", "#33CC33"),
          cex = 0.8,
          fontfamily = "serif",
          cat.cex = 0.8,
          cat.pos = c(-0, 0),
          cat.dist = c(0.03, 0.03),
          cat.fontfamily = "serif",
          disable.logging = TRUE)

venn.diagram(x = list(KO_Data_Venn_Down,GCGR_Ab_Data_Venn_Down),
             category.names = c(expression("GCGR"^"-/-") , "GCGR Ab"),
             filename = 'VennDiagram_GCGR_KO_Ab_DOWN_LIPID.png',
             output=TRUE,
             main = "Down-regulated genes",
             main.pos = c(0.5,0.95),
             main.cex = 0.9,
          imagetype="png",
          resolution = 1000,
          compression = "lzw",
          lty = "blank",
          fill = c("#330099", "#33CC33"),
          cex = 0.8,
          fontfamily = "serif",
          cat.cex = 0.8,
          cat.pos = c(-0, 0),
          cat.dist = c(0.03, 0.03),
          cat.fontfamily = "serif",
          disable.logging = TRUE)

Down_Overlap <- calculate.overlap(x = list(KO_Data_Venn_Down,GCGR_Ab_Data_Venn_Down)) %>% .[["a3"]] %>% as.data.frame(row.names = "ENSEMBL") %>% t() %>% as_tibble() %>% semi_join(All_Sig_Genes_GCGR_Ab,., by = "ENSEMBL") %>% dplyr::select("ENSEMBL","gene_name","gene_symbol","padj","foldChange")

Up_Overlap <- calculate.overlap(x = list(KO_Data_Venn_Up,GCGR_Ab_Data_Venn_Up)) %>% .[["a3"]] %>% as.data.frame(row.names = "ENSEMBL") %>% t() %>% as_tibble() %>% semi_join(All_Sig_Genes_GCGR_Ab,., by = "ENSEMBL") %>% dplyr::select("ENSEMBL","gene_name","gene_symbol", "padj","foldChange")

All_Sig_Genes_KOvsWT_Overlap <- rename(All_Sig_Genes_KOvsWT, foldChange_KO = foldChange, padj_KO = padj) %>% dplyr::select("ENSEMBL", "padj_KO","foldChange_KO")

Down_Overlap <- left_join(Down_Overlap, All_Sig_Genes_KOvsWT_Overlap, by = "ENSEMBL") %>% rename(., foldChange_Ab = foldChange, padj_Ab = padj)
Up_Overlap <- left_join(Up_Overlap, All_Sig_Genes_KOvsWT_Overlap, by = "ENSEMBL") %>% rename(., foldChange_Ab = foldChange, padj_Ab = padj)

write_xlsx(Down_Overlap, "Overlapping_Down_Genes_GcgrKO_GCGR-Ab_LIPID.xlsx")
write_xlsx(Up_Overlap, "Overlapping_UP_Genes_GcgrKO_GCGR-Ab_LIPID.xlsx")
```

Creating a list of amino acid annotated genes (GOBP annotated): 
```{r message=FALSE, warning=FALSE}
AA_GOBP <- unique(c(GOBPOFFSPRING[["GO:0006520"]],GOBPOFFSPRING[["GO:0006865"]],GOBPOFFSPRING[["GO:0080144"]], GOBPOFFSPRING[["GO:0043200"]]))

AA_BP_Genes <- bitr(AA_GOBP, fromType="GOALL", toType= "SYMBOL", OrgDb='org.Mm.eg.db')
AA_BP_Genes <- tibble("gene_symbol" = unique(AA_BP_Genes$SYMBOL))
```
  
Comparing downregulated amino acid related genes in the two datasets and exporting a list of the overlapping genes:
```{r eval=FALSE, include=TRUE}
KO_Data_Venn_Down <- All_Sig_Genes_KOvsWT %>% semi_join(., AA_BP_Genes, by = "gene_symbol") %>%
  filter(., log2FoldChange < 0) %>% 
  dplyr::select('ENSEMBL') %>%
  c(unlist(.['ENSEMBL'])) %>% .[-1]
GCGR_Ab_Data_Venn_Down <- All_Sig_Genes_GCGR_Ab %>% semi_join(., AA_BP_Genes, by = "gene_symbol") %>%
  filter(., log2FoldChange < 0) %>% 
  dplyr::select('ENSEMBL') %>%
  c(unlist(.['ENSEMBL'])) %>% .[-1]

venn.diagram(x = list(KO_Data_Venn_Down,GCGR_Ab_Data_Venn_Down),
             category.names = c(expression("GCGR"^"-/-") , "GCGR Ab"),
             filename = 'VennDiagram_GCGR_KO_Ab_DOWN_AA.png',
             output=TRUE,
             main = "Down-regulated genes",
             main.pos = c(0.5,0.95),
             main.cex = 0.9,
          imagetype="png",
          resolution = 1000,
          compression = "lzw",
          lty = "blank",
          fill = c("#330099", "#33CC33"),
          cex = 0.8,
          fontfamily = "serif",
          cat.cex = 0.8,
          cat.pos = c(10, -10),
          cat.dist = c(0.03, 0.03),
          cat.fontfamily = "serif",
          disable.logging = TRUE)

Down_Overlap <- calculate.overlap(x = list(KO_Data_Venn_Down,GCGR_Ab_Data_Venn_Down)) %>% .[["a3"]] %>% as.data.frame(row.names = "ENSEMBL") %>% t() %>% as_tibble() %>% semi_join(All_Sig_Genes_GCGR_Ab,., by = "ENSEMBL") %>% dplyr::select("ENSEMBL","gene_name","gene_symbol","padj","foldChange")

All_Sig_Genes_KOvsWT_Overlap <- rename(All_Sig_Genes_KOvsWT, foldChange_KO = foldChange, padj_KO = padj) %>% dplyr::select("ENSEMBL", "padj_KO","foldChange_KO")

Down_Overlap <- left_join(Down_Overlap, All_Sig_Genes_KOvsWT_Overlap, by = "ENSEMBL") %>% rename(., foldChange_Ab = foldChange, padj_Ab = padj)

write_xlsx(Down_Overlap, "Overlapping_Down_Genes_GcgrKO_GCGR-Ab_AA.xlsx")
```

### Session Info

```{r}
sessionInfo()
```